Appendix S1

### 2020-06-11 analysis/paper/appendix_s1.Rmd

One of the crucial steps in building this carrying capacity model was choosing which habitat covariates to include. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points). Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g. substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011]. MIC is a measure of correlation that incorporates potential non-linear associations, so for example if there is an quadratic association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g. meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value of each measured habitat characteristics and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by restoration actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many all zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        format = 'html',
        # format = 'latex',
        booktabs = T,
        caption = 'MIC statistic for top metrics in each habitat category. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T)

@@

```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   rownames(x) <- tibble(Metric = rownames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   
  #   # colnames(x) <- tibble(Metric = colnames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   return(x)
  # }) %>%
    map(.f = corrplot::corrplot.mixed,
        upper = 'ellipse',
        tl.pos = 'lt',
        order = 'FPC')
```

```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, facted by habitat category."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

```{r selected-corr, fig.cap = 'Correlation plot of habitat metrics used in QRF model.'}
hab_data %>% 
  select(one_of(sel_hab_mets$Metric)) %>% 
  cor(use = 'pairwise',
      method = 'pearson') %>%
      # method = 'spearman') %>% 
  # as_cordf() %>%
  ggcorrplot::ggcorrplot(method = 'circle',
                         # hc.order = T,
                         type = 'lower',
                         # legend.title = "Correlation",
                         colors = c('blue4', 'white', 'red3'))
```

>>>>

One of the crucial steps in building this carrying capacity model was choosing which habitat covariates to include. Random forest models naturally incorporate interactions between correlated covariates, which is essential since nearly all habitat variables are considered correlated to one degree or another, however, we aimed to avoid overly redundant variables (i.e., variables that measure similar aspects of the habitat). Further, including too many covariates can result in overfitting of the model (e.g., including as many covariates as data points). Our goal was to select a group of covariates that captured as many different aspects of the stream habitat (e.g., substrate, flow, riparian condition, channel unit configuration, etc.) as possible, while still holding information about fish densities.

To prevent overfitting, we pared down the more than 100 metrics generated by the CHaMP protocol describing the quantity and quality of fish habitat for each survey site. Habitat metrics were first grouped into broad categories that included channel unit configuration, complexity, fish cover, riparian areas, side channels, stream size, substrate, temperature, water quality, and woody debris. Habitat metrics measuring any large wood volume were scaled by the site length (in 100 m units). To assist in determining the habitat metrics to include in the QRF model, we used the Maximal Information-Based Nonparametric Exploration (MINE) class of statistics [@Reshef2011] to determine those habitat characteristics (covariates) most highly associated with the log of observed parr densities. We calculated the maximal information coefficient (MIC), using the R package *minerva* [@R-minerva], to measure the strength of the linear or non-linear association between the natural log of fish density and each habitat metric [@Reshef2011]. MIC is a measure of correlation that incorporates potential non-linear associations; for example, if there is a quadratic association the MIC value could be high, even when the standard correlation coefficient is low. We excluded categorical variables such as channel type (e.g., meandering, pool-riffle, plane-bed, etc.) because we assumed that other quantitative metrics would capture the differences between those qualitative categorical metrics.

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by rehabilitation actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        format = 'html',
        # format = 'latex',
        booktabs = T,
        caption = 'MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T)

@@

```{r corr-fig1, eval = F, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)] %>%
  # map(.f = function(x) {
  #   rownames(x) <- tibble(Metric = rownames(x)) %>%
  #     left_join(hab_dict %>%
  #                 select(Metric = ShortName,
  #                        Name) %>%
  #                 distinct()) %>%
  #     pull(Name)
  #   
  #   # colnames(x) <- tibble(Metric = colnames(x)) %>%
  #   #   left_join(hab_dict %>%
  #   #               select(Metric = ShortName,
  #   #                      Name) %>%
  #   #               distinct()) %>%
  #   #   pull(Name)
  #   return(x)
  # }) %>%
    map(.f = corrplot::corrplot.mixed,
        upper = 'ellipse',
        tl.pos = 'lt',
        order = 'FPC')
```

```{r corr-fig, results='hide', fig.width = 5, fig.height = 5, fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

```{r selected-corr, fig.cap = 'Correlation plot of habitat metrics used in the QRF model.'}
hab_data %>% 
  select(one_of(sel_hab_mets$Metric)) %>% 
  cor(use = 'pairwise',
      method = 'pearson') %>%
      # method = 'spearman') %>% 
  # as_cordf() %>%
  ggcorrplot::ggcorrplot(method = 'circle',
                         # hc.order = T,
                         type = 'lower',
                         # legend.title = "Correlation",
                         colors = c('blue4', 'white', 'red3'))
```
